---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# gammabuster

The gammabuster package contains two functions that translate judgmental estimates about a random variable into the gamma distribution parameters consistent with that judgement. This is helpful because the parameters of a gamma distribution--the shape and scale parameters--don't have a real-world interpretation. This makes it hard to elicit judgmental estimates of a gamma distribution via its parameters directly. However, it's relatively straightforward to elicit judgments about the likely upper and lower bounds of a distribution or about the mean and range of the distribution. The gammabuster package solves this problem. It converts judgmental estimates about a gamma distribution into needed parameter estimates.

## Installation

You can install gammabuster from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("JohnsonBrent/gammabuster")
```

## Example

Here's one example. Suppose one has elicited judgments that the range of a random variable lies between 0 and 5. Assuming the random variable is strictly positive and continuous within this space, a gamma might be the appropriate distribution. If so, the following function generates the gamma parameters consistent with the judgement:

```{r example, echo = TRUE, eval=FALSE}
GbusterMinMax(0, 5)
```

There are more examples and a deeper explanation in the vignette. 

